---
title: "16S_analysis"
author: "Tristan Caro"
date: "8/30/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries

```{r, message=FALSE}
library(mctoolsr)
library(plyr) # For hulls
library(tidyverse) # For data manipulation and plotting
library(RColorBrewer) # For color palettes
library(vegan) # For alpha and beta diversity analyses
library(indicspecies) # For indicator species
library(car) # For Levene Test and anova
library(PMCMR) # For Nemenyi posthoc test
```

# Startup

## Import Tax and Map Files

```{r}
tax_table_fp = 'seqtab_wTax_mctoolsr_16s_caro.txt'
map_fp = 'caro_16s_mapping.txt' 
input = load_taxa_table(tax_table_fp, map_fp)
```
## Remove unwanted sequences
Don't want mito, chloro, eukaryote, or domain-unassigned seqs.
> Get those guys out of here!

```{r}
# Remove mitochondrial and chloroplast sequences, Remove reads assigned as eukaryotes
input_filt <- filter_taxa_from_input(input, taxa_to_remove = c("Chloroplast","Mitochondria", "Eukaryota"))
# Remove reads that are unassigned at domain level
input_filt <- filter_taxa_from_input(input_filt, at_spec_level = 2, taxa_to_remove = "NA")

```

We now have an object called input. It contains 3 files - a sequence table, a mapping file, and a taxonomic file.

```{r}
head(input$data_loaded) # Rows are OTUs, columns are samples
```

```{r}
head(input$taxonomy_loaded) # Rows are ASVs, columns are taxonomic levels
```

```{r}
head(input$map_loaded) # Rows are samples, columns are variables

```

# Initial data examination, rarefaction
One of the first things we want to do is see how many sequences per sample there are. This is done by getting the column sums of the sequence table, and we'll sort it too.

```{r}
sort(colSums(input$data_loaded))
```

We could also save this as an object and plot it
```{r}
seqcounts <- as.data.frame(sort(colSums(input$data_loaded)))
```

This puts the sample names as row names and the sequence counts column is titled `sort(colSums(input$data_loaded)))`. We'll use the pipeline %>% which is very useful for dataframe management with the dplyr package. We'll rename the column and make a new column all at once.

```{r}
seqcounts <- as.data.frame(sort(colSums(input$data_loaded))) %>%
  rename("seqs" = "sort(colSums(input$data_loaded))") %>%
  rownames_to_column(var = "sampleID")
```

Now we have a better dataframe with two columns, seqs and sampleID which we can plot!

```{r}

p_seqcounts <- seqcounts %>% 
  ggplot(aes(reorder(sampleID, seqs, mean), seqs)) + # Dataframe and variables
    geom_col() +
    labs(y = "# Reads", x = "Sample") +
    coord_flip() + # Flip axes
    theme_classic() # Plot style
p_seqcounts
```
This shows that we have anywhere from 27,000-60,000 reads per sample

## Rarefaction
Moving on, let's rarefy the data at the lowest count per sample. Sometimes you may want to drop samples and choose a higher number.
Rarefy at 27,000 seqs/sample.

```{r}
input_rar = single_rarefy(input, 27000) # This makes a new mctoolsr dataset called input_rar. The blank sample dropped out!
```
Check seq counts in new dataset
```{r}
colSums(input_rar$data_loaded) 
```
Good - all 27000!

> Note: normally after you've rarefied and filtered it's recommended to save the dataset as a .rds file. This insures you use the same dataset and analysis is reproducible (because rarefaction involves random sampling). Then, if you quit your session but want to reanalyze something, you would ignore the rarefaction/filtering and just start right here!

```{r}
saveRDS(input_rar, file = "input_rar.rds")
input_rar <- readRDS("input_rar.rds")
```

# Alpha Diversity
Not extremely useful, but people report it.
Let's analyze and graph the number of ASVs in our sample types
Two categories: t-test (parametric) or Wilcoxon Test (non-parametric)
Three + categories: ANOVA (parametric) or Kruskal-Wallis Test (non-parametric)

First, let's get the number of ASVs (richness) per sample and add it to the mapping file.
```{r}
input_rar$map_loaded$rich <- specnumber(input_rar$data_loaded, MARGIN = 2)
```
>Note: since the data_loaded file has ASVs as rows, MARGIN must be set to 2.

```{r}
# Three or more categories
leveneTest(input_rar$map_loaded$rich ~ input_rar$map_loaded$combo_label)
# Variance homogeneous (p > 0.05)

m <- aov(input_rar$map_loaded$rich ~ input_rar$map_loaded$combo_label)
shapiro.test(m$residuals)
# Residuals normally distributed (p > 0.05)
```
```{r}
# Other diagnostics - learn more here (https://data.library.virginia.edu/diagnostic-plots/)
plot(m) # Click in the console and hit Return to see each diagnostic plot
summary(m)
TukeyHSD(m)
```

```{r}
p_alpha_div <- input_rar$map_loaded %>% ggplot(aes(combo_label, rich, color = combo_label)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(size = 3, 
              alpha = 0.5) +
  labs(x = "Soil Type", 
       y = "ASV Richness", 
       colour = "combo_label",
       title = "Alpha Diversity") +
  scale_colour_viridis_d() +
  theme_bw() +
  theme(axis.title = element_text(face = "bold",
                                  size = 16), 
        axis.text = element_text(size = 14),
        legend.position = "none")
p_alpha_div
```

# Beta Diversity
Let's calculate Bray-Curtis dissimilarity, plot some PCoA ordinations, and run some stats.

## Bray-Curtis Matrix
```{r}
bc <- calc_dm(input_rar$data_loaded)
```

## Principle Coordinate Analysis (PCoA)
```{r}
pcoa <- cmdscale(bc, k = nrow(input_rar$map_loaded) - 1, eig = T)
```

## Variation Explained
```{r}
eigenvals(pcoa)/sum(eigenvals(pcoa)) # 25.7, 18.9 % variation explained
```

```{r}
# Save Axis 1 and 2 to the mapping file
input_rar$map_loaded$Axis01 <- scores(pcoa)[,1]
input_rar$map_loaded$Axis02 <- scores(pcoa)[,2]

# Function for making a convex hull
find_hull <- function(df) df[chull(df$Axis01, df$Axis02),]

# Calculate hulls and save to dataframe
micro.hulls <- ddply(input_rar$map_loaded, c("combo_label"), find_hull)
```

```{r}
p_pcoa <- input_rar$map_loaded %>% ggplot(
  aes(Axis01, Axis02, color = combo_label)) +
  geom_polygon(data = micro.hulls, 
               aes(colour = combo_label),
               alpha = 0.1, 
               show.legend = F) +
  geom_point(size = 2, 
             alpha = 0.5) +
  labs(x = "PC1: 25.7% Variation Explained", 
       y = "PC2: 18.9% Variation Explained",
       colour = "Soil Type") +
  scale_colour_viridis_d() +
  scale_fill_viridis_d() +
  theme_bw() +  
  theme(legend.position = "right",
        axis.title = element_text(face = "bold", 
                                  size = 16), 
        axis.text = element_text(size = 14))
p_pcoa
```

## PERMANOVA
Test for differences in centroid of different PCoA hull groups.
```{r}
set.seed(1223) # To make reproducible 
adonis(bc ~ input_rar$map_loaded$combo_label)
```
Significant effects of type.

## PERMDISP
Multivariate version of Levenne Test. Difference in variation in each factor level
```{r}
m1 <- betadisper(bc, input_rar$map_loaded$combo_label)
anova(m1) # Dispersion homogeneous for soil type
```

# Taxonomic Analyses
Indicator taxa - SIMPER (Similarity Percentages) or MULTIPATT
both take into account the number of samples that the taxa is present in within each group as well as abundance. SIMPER (list how much each ASV contributes to dissimilarity among groups).

## SIMPER

```{r}
sim <- simper(t(input_rar$data_loaded), 
              input_rar$map_loaded$combo_label)
s <- summary(sim)
```

Let's look at the top 5 contributing to dissimilarity between the two types with significantly different soils
```{r}
head(s$GGm0_GGm3, n = 5)
```
Average is the proportion contribution, cumsum is cumulative, ava and avb are mean sequence abundances per group.

## MULTIPATT

MULTIPATT (list ASVs associated with each group)
Difference between this and SIMPER is in what it is reporting:
MULTIPATT doesn't give a value for every ASV It only outputs important ASVs.

**Excluded for brevity.**

```
set.seed(1223) # For reproducibility
mp <- multipatt(t(input_rar$data_loaded), 
                input_rar$map_loaded$combo_label, 
                func = "IndVal.g", 
                control = how(nperm=999))
summary(mp)
```

## Main taxa at different levels

Let's quickly just use the mctoolsr base functions
Summarize at family level. Choose different levels with the level argument

```{r}
tax_sum_families <- summarize_taxonomy(input_rar, level = 5, report_higher_tax = FALSE) #level 5 = families, level 6 = genus
```

### Heatmap
mctoolsr Default heatmap, which uses ggplot2 in the background so we can use normal ggplot2 syntax numbers are percent relative abundances

```{r}
plot_ts_heatmap(tax_sum_families, 
                input_rar$map_loaded, 
                0.01, 
                'combo_label',
                rev_taxa = T) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1))
```
### Stacked Bar Chart

```{r}
plot_taxa_bars(tax_sum_families,
               input_rar$map_loaded,
               "combo_label",
               num_taxa = 10) +
  labs(x = "Soil Type", y = "Relative Abundance", fill = "Family") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))
```

```{r}
bars <- plot_taxa_bars(tax_sum_families,
                       input_rar$map_loaded,
                       "combo_label",
                       num_taxa = 10,
                       data_only = TRUE) # allows us to save the data and not make the plot
ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", colour = "black", size = 0.25) +
  labs(x = "Soil Type", y = "Relative Abundance", fill = "Family") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))
```

### Kruskal-Wallis Test

Test (run a Kruskal-Wallis test on all families with mean rel abund. > filter_level in at least one of the factor levels) which families are different between the sample types (testing all at once)

```{r}
taxa_summary_by_sample_type(tax_sum_families, 
                            input_rar$map_loaded, 
                            type_header = 'combo_label', 
                            filter_level = 0.05, # cut off for relative abundance
                            test_type = 'KW') # Bonferroni corection imporant since you did many tests at once

```
```{r}
tax_sum_phyla <- summarize_taxonomy(input_rar, level = 2, report_higher_tax = FALSE) #level 5 = families, level 6 = genus
```

```{r}
# Add some customization
plot_ts_heatmap(tax_sum_phyla, 
                input_rar$map_loaded, 
                0.01, 
                'combo_label',
                rev_taxa = T) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1))
```
```{r}
# Add some customization
plot_taxa_bars(tax_sum_phyla,
               input_rar$map_loaded,
               "combo_label",
               num_taxa = 10) +
  labs(x = "Soil Type", y = "Relative Abundance", fill = "Phyla") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))
```
```{r}
bars <- plot_taxa_bars(tax_sum_phyla,
                       input_rar$map_loaded,
                       "combo_label",
                       num_taxa = 10,
                       data_only = TRUE) # allows us to save the data and not make the plot

ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", colour = "black", size = 0.25) +
  labs(x = "Soil Type", y = "Relative Abundance", fill = "Phyla") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14))
```
```{r}
# Test (run a Kruskal-Wallis test on all families with mean rel abund. > filter_level in at least one of the factor levels)
# which phyla are different between the sample types (testing all at once)
taxa_summary_by_sample_type(tax_sum_phyla, 
                            input_rar$map_loaded, 
                            type_header = 'combo_label', 
                            filter_level = 0.05, # cut off for relative abundance
                            test_type = 'KW') # Bonferroni correction important since you did many tests at once

```

Verrucomicrobia is significantly different!

# Variation between replicates - graphing individual samples

Main taxa at different levels
Let's quickly just use the mctoolsr base functions
Summarize at family level. Choose different levels with the level argument

```{r}
tax_sum_phyla <- summarize_taxonomy(input_rar, level = 2, report_higher_tax = FALSE) #level 5 = families, level 6 = genus

```

```{r}
plot_ts_heatmap(tax_sum_phyla, 
                input_rar$map_loaded, 
                0.01, 
                'names',
                rev_taxa = T) +
  coord_flip() +
  theme(axis.text.x = element_text(size = 12, angle = 45, vjust = 1))
```

```{r}
bars <- plot_taxa_bars(tax_sum_phyla,
                       input_rar$map_loaded,
                       "names",
                       num_taxa = 11,
                       data_only = TRUE) # allows us to save the data and not make the plot
ggplot(bars, aes(group_by, mean_value, fill = taxon)) +
  geom_bar(stat = "identity", colour = "black", size = 0.25) +
  labs(x = "Soil Type", y = "Relative Abundance", fill = "Phyla") +
  scale_fill_brewer(palette = "Paired") +
  theme_bw() +
  theme(axis.title = element_text(face = "bold", size = 16), 
        axis.text = element_text(size = 14, angle = 90))
```
Looks good to me.
